---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## load necessary packages, installing if necessary
```{r}
if (!"ggplot2" %in% installed.packages()) install.packages("ggplot2")
library("ggplot2")

if (!"plyr" %in% installed.packages()) install.packages("plyr")
library("plyr")
```


## Loading and preprocessing the data
```{r}
rawData <- read.csv(unz("activity.zip", "activity.csv"), header=TRUE, sep=",")

```


## What is mean total number of steps taken per day?
Total number of steps taken per day
```{r}
dailyTotals <- ddply(rawData, c("date"), summarise, sum = sum(steps, na.rm=TRUE))

dailyTotals
```

Make a histogram of the total number of steps taken each day
```{r}
qplot(dailyTotals$sum,
      geom="histogram",
      main="Total number of Steps per Day Histogram",
      xlab = "Steps per Day",
      fill=I("blue"))
```

Mean of total number of steps taken per day
```{r}
mean(dailyTotals$sum, na.rm=TRUE)

```

Median of total number of steps taken per day
```{r}
median(dailyTotals$sum, na.rm=TRUE)

```


## What is the average daily activity pattern?
Average steps taken during 5-minute interval over all days
```{r}
intervalAvg <- ddply(rawData, c("interval"), summarise, mean=mean(steps, na.rm=TRUE))

intervalAvg

```

Time series plot of 5-minute interval and average steps taken
```{r}
ggplot(intervalAvg, aes(x=interval, y=mean)) + 
      geom_line() +
      xlab("Interval") +
      ylab("Average over Days")
```

The 5-minute interval with the highest average steps taken
```{r}
intervalAvg[which.max(intervalAvg$mean),]
```

## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
